Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation

### 2021-08-24 summary.tex
Premerge

\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\usepackage{mathtools}
% \usepackage{natbib}
\usepackage[backend=bibtex,style=authoryear,giveninits,natbib]{biblatex}
\addbibresource{misc.bib}
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}
\usepackage{parskip}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
% \setcitestyle{numbers,square,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\title{Popularity Adjusted Block Models are Generalized Random Dot Product Graphs}

\author{John Koo\\Department of Statistics, Indiana University\\
and\\
Minh Tang\\Department of Statistics, North Carolina State University\\
and\\
Michael Trosset\\Department of Statistics, Indiana University}

% \date{\vspace{-2.5em}}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We connect two random graph models, the Popularity Adjusted Block Model
(PABM) and the Generalized Random Dot Product Graph (GRDPG),
by demonstrating that a PABM is a GRDPG in which communities correspond to 
mutually orthogonal subspaces of latent vectors. This insight
leads to the construction of improved algorithms for community detection
and parameter estimation with PABM. Using established asymptotic
properties of Adjacency Spectral Embedding for GRDPG, we derive
asymptotic properties of these algorithms, including algorithms that
rely on Sparse Subspace Clustering. We illustrate these properties
via simulation study.
\end{abstract}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newpage
\spacingset{1.5} % DON'T change the spacing!

@@

Statistical inference on random graphs requires a suitable probability model.  A general probability model for unweighted and undirected graphs is the Bernoulli Graph, which assumes that edges occur as independent Bernoulli trials.  A Bernoulli Graph is characterized by an edge probability matrix $P=[P_{ij}]$, where an edge between vertices $i$ and $j$ occurs with success probability $P_{ij}$.  A trivial example of a Bernoulli Graph is the Erd\"{o}s-R\'{e}nyi model proposed by \citet{Gilbert:1959}, in which the vertices of the random graph are fixed and possible edges occur independently with fixed probability $P_{ij}=p$.  The requirement that $p$ is fixed is too strong for most applications, and various researchers have weakened that requirement in various ways.  The present work relates two lines of generalization.

Network analysis is often concerned with community detection.  One form of community detection assumes that each vertex belongs to an unobserved community, with the probability of an edge between vertices $i$ and $j$ depending on the communities to which $i$ and $j$ belong.  Formally, one assigns each vertex $v_i$ a community label $z_i$ and assumes a Bernoulli Graph in which $P_{ij}$ is a function of $z_i$ and $z_j$.  Such models, called Block Models, define the goal of community detection as a problem in statistical inference: identify the true community (up to permutation of labels) to which each vertex belongs.

The classical Stochastic Block Model (SBM) of \citet{doi:10.1080/0022250X.1971.9989788} specifies that each edge probability $P_{ij}$ depends only on the labels $z_i$ and $z_j$, i.e., $P_{ij} = \omega_{z_,z_j}$.  Subsequent researchers have weakened this assumption.   The Degree-Corrected Block Model (DCBM) of \citet{Karrer_2011} assigns an additional parameter $\theta_i$ to each vertex and sets $P_{ij} = \theta_i \theta_j \omega_{z_,z_j}$. The Popularity Adjusted Block Model (PABM) of \citet{307cbeb9b1be48299388437423d94bf1} generalizes the DCBM, allowing heterogeneity of edge probabilities within and between communities while still maintaining distinct community structure.

Another type of Bernoulli Graph was proposed by \citet*{10.1007/978-3-540-77004-6_11}.  A Random Dot Product Graph (RDPG) specifies that each vertex corresponds to a latent position vector in Euclidean space and that the probability of an edge between two vertices is the dot product of their latent position vectors.  Thus, if the latent positions $x_1,\ldots,x_n \in \mathbb{R}^d$ and $X = [\begin{array}{ccc} x_1 & \cdots & x_n \end{array} ]^\top$, then the edge probability matrix is $P = XX^\top$.  Clearly, any Bernoulli Graph with positive semidefinite $P$ is an RDPG.  The positive definite Euclidean inner product in the RDPG model was replaced by an indefinite inner product in \citet{rubindelanchy2017consistency}, resulting in the {\em Generalized}\/ RDPG (GRDPG).

In contrast to Block Models, neither RDPGs nor GRDPGs inherently specify distinct communities.  However, one can easily impose community structure by assuming that the latent positions lie in distinct clusters.  Hence, it is not surprising that Block Models can be studied by reformulating them as RDPGs or GRDPGs.  For example, an assortative SBM (an SBM for which $P$ is positive semidefinite) is equivalent to an RDPG for which all vertices in the same community correspond to the same latent position vector (\cite{lyzinski2014}).  Likewise, the DCBM is equivalent to an RDPG for which all vertices in the same community correspond to latent position vectors that lie on a straight line (\cite{lyzinski2014, rubindelanchy2017consistency}).

Because the edge probability matrix of a PABM is not necessarily positive semidefinite, a PABM is not necessarily an RDPG.   In Section \ref{connecting-the-pabm-to-the-grdpg} we demonstrate that every PABM is in fact a specific type of GRDPG for which the latent position vectors lie in distinct orthogonal subspaces, each subspace corresponding to a community.  This identification is our central result.  In Section \ref{methods}, we use the geometry of the GRDPG to derive more efficient algorithms for detecting the communities and estimating the parameters in the PABM.  We report the results of simulation studies in Section \ref{simulated-examples} and apply our methods to three well-known data sets in Section \ref{real-data-examples}. Section \ref{discussion} concludes.

Statistical inference on random graphs involves the
partitioning of a graph into disconnected subgraphs or clusters. This is
often motivated by the assumption that there exist underlying and
unobserved communities to which each vertex of the graph belongs, and
edges between pairs of vertices are determined by drawing from a
probability distribution based on the community relationships between
each pair. The goal of this analysis then is population community
detection, or the recovery of the true underlying community labels for
each vertex, up to permutation, with some additional parameter
estimation being of possible interest, assuming some underlying
probability model.

A general family of models that describe unweighted and undirected graphs is 
the Bernoulli Graph, which is characterized by edge probability matrix $P$. 
Edges between each pair of vertices $i$ and $j$ are drawn 
as a Bernoulli trial with probability $P_{ij}$, for each 
$0 \leq i < j \leq n$ and $n$ is the number of vertices in the graph. 
Community detection methods often restrict this type of model 
by assigning each vertex a community label $z_1, ..., z_n$ 
and then setting each $P_{ij}$ as a function of the labels $z_i$ and $z_j$. 
Such models are called Block Models.

One popular type of Block Model is the Stochastic Block Model (SBM),
first proposed by \citet{doi:10.1080/0022250X.1971.9989788}, 
which restricts each edge probability $P_{ij}$ as only depending on the labels 
$z_i$ and $z_j$, i.e., $P_{ij} = \omega_{z_i, z_j}$. 
Various generalizations of the SBM have since been utilized, 
including the Degree-Corrected Block Model (DCBM), 
introduced by \citet{Karrer_2011}, 
which assigns an additional parameter $\theta_i$ to each vertex 
and sets $P_{ij} = \theta_i \theta_j \omega_{z_i, z_j}$. 
The Popularity Adjusted Block Model (PABM) was then introduced by
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the
DCBM to address the heterogeneity of edge probabilities within and
between communities while still maintaining distinct community
structure.

Parallel to the families of Block Models is another type of Bernoulli Graph 
called the Random Dot Product Graph (RDPG), proposed by
\citet*{10.1007/978-3-540-77004-6_11}. 
Under this model, each vertex of the graph
can be represented by a vector in some latent space such that the edge
probability between any pair of vertices is given by their corresponding
dot product in the latent space, i.e., given a latent positions
\(x_1, ..., x_n \in \mathbb{R}^d\), the edge probability matrix is
\(P = X X^\top\) where 
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\). 
It can be shown that all Bernoulli Graphs with 
positive semidefinite edge probability matrix $P$ is a type of RDPG. 
For instance, the assortative
SBM is equivalent to a special case of the RDPG for which all vertices
of a given community share the same position in the latent space
\cite{lyzinski2014}. A similar property exists for the DCBM, 
which is equivalent to a RDPG in which vertices that belong to the same 
community all lie in a line segment in the latent space 
\cite{lyzinski2014, rubindelanchy2017consistency}. 
Because the edge probability matrix for the PABM is not positive semidefinite, 
we instead show that it is a special case of the 
\emph{Generalized} Random Dot Product Graph (GRDPG) 
in order to analyze the PABM as a latent space model 
for community detection and parameter estimation.

\hypertarget{connecting-the-popularity-adjusted-block-model-to-the-generalized-random-dot-product-graph}{%
\section{PABMs are GRDPGs}\label{connecting-the-popularity-adjusted-block-model-to-the-generalized-random-dot-product-graph}}

In this section, we show that the PABM is a special case of the GRDPG. 
More specifically, graph $G$ drawn from the PABM can be represented by 
a collection of latent vectors in Euclidean space. 
We further show that the latent configuration that
induces the PABM consists of orthogonal subspaces with each subspace
corresponding to a community.

@@

Let $G = (V, E)$ be an unweighted, undirected, 
and hollow graph with vertex set $V$ ($|V| = n$) and edge set $E$. 
$A \in \{0, 1\}^{n \times n}$ represents the adjacency matrix of $G$ 
such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ 
and $0$ otherwise. Because $G$ is symmetric and hollow,
$A_{ij} = A_{ji}$ and $A_{ii} = 0$ for each $i, j \in [n]$. 
We further restrict our analyses to Bernoulli graphs. 
Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. 
Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ 
for each $1 \leq i < j \leq n$ (setting $A_{ji} = A_{ij}$ and $A_{ii} = 0$). 
We denote $A \sim \BG(P)$ as a graph with adjacency matrix $A$ 
sampled from edge probability matrix $P$ in this manner. 
If each vertex has a hidden label in $[K]$, 
they are denoted as $z_1, ..., z_n$. 
Finally, we denote 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ 
as the collection of $n$ latent vectors $x_1, ..., x_n \in \mathbb{R}^d$.

\hypertarget{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}{%
\subsection{Two Probability Models for Graphs}\label{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}}

@@

Another characterization of the PABM is as follows.
Let the rows and columns of $P$ be arranged by community label
such that $n_k \times n_{\ell}$ block $P^{(k \ell)}$
describes the edge probabilities between vertices in communities
$k$ and $\ell$; note that $P^{(k \ell)} = (P^{(k \ell)})^\top$.
Now suppose that each block $P^{(k \ell)}$ can be written as the outer product of two vectors:

\begin{equation} \label{eq:pabm}
  P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top
\end{equation}

for a set of $K^2$ popularity vectors $\{\lambda^{(st)}\}_{1 \leq
  s \leq K, 1 \leq t \leq K}$ where each
$\lambda^{(st)}$ is a column vector of dimension $n_s$. Then a graph $G$
is drawn from a PABM with parameters
$\{\lambda^{(st)}\}$ if its adjacency matrix $A$
satisfies $A \sim \mathrm{Bernoulli}(P)$.
\end{definition}

We will use the notation \(A \sim \PABM(\{\lambda^{(k \ell)}\}_K)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with parameters
\(\{\lambda^{(k \ell)}\}\) consisting of \(K\) underlying communities.

\begin{definition}[Generalized Random Dot Product Graph]
\label{grdpg}
Let graph $G = (V, E)$ be drawn as $A \sim \BG(P)$.
If there exists a data matrix $X \in \mathbb{R}^{n \times d}$ such that

@@

\begin{remark}
We can use Adjacency Spectral Embedding (ASE) 
\cite{rubindelanchy2017statistical} to recover the latent vectors of a GRDPG. 
This procedure consists of taking the spectral decomposition of $A$ 
(or $P$ if available) and choosing the $p$ most positive and $q$ most negative
eigenvalues and their corresponding eigenvectors 
to construct the embedding $\hat{Z} = V |D|^{1/2}$. 
\end{remark}

@@

\begin{remark}
The latent vectors that produce $X I_{p,q} X^\top = P$ are not unique
\cite{rubindelanchy2017statistical}.
More specifically, if $P_{ij} = x_i^\top I_{p, q} x_j$, 
then we also have for any$Q \in \mathbb{O}(p, q)$,
$(Q x_i)^\top I_{p, q} (Q x_j) = x_i^\top (Q^\top I_{p, q} Q) x_j =
x_i^\top I_{p, q} x_j = P_{ij}$.
Unlike in the RDPG case, transforming the latent positions via multiplication
by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or
distances.
\end{remark}

\hypertarget{connecting-the-pabm-to-the-grdpg}{%
\subsection{The Geometry of PABMs}\label{connecting-the-pabm-to-the-grdpg}}

Now that we defined the PABM and GRDPG, 
we show the special geometry of the PABM when viewed as a GRDPG.

\begin{theorem}[The latent configuration of the PABM]
\label{theorem2}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\}_K)$ be an instance of a
PABM with $K \geq 1$ blocks and latent vectors $\{\lambda^{(k \ell)}
\colon 1 \leq k \leq K, 1 \leq \ell \leq K\}$. 
Then there exists a block diagonal matrix
$X \in \mathbb{R}^{n \times K^2}$ defined by $\{\lambda^{(k \ell)}\}$ and a 
$K^2 \times K^2$ {\em fixed} orthonormal matrix $U$ such 
that $A \sim \mathrm{GRDPG}_{K (K+1) / 2, K (K-1) / 2}(XU)$.
\end{theorem}

\begin{proof}
We will prove this theorem in two parts. First, for demonstration purposes, we focus on the case for $K = 2$. Then we generalize this to $K \geq 2$.

For the $K = 2$ case, the proof is straightforward. Let

$$X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
0 & 0 & \lambda^{(21)} & \lambda^{(22)}
\end{bmatrix} \quad \text{and} \quad
U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
0 & 1 & 0 & 0 \end{bmatrix}.$$

Then by straightforward matrix multiplication, we obtain

\[X U I_{3, 1} U^\top X^\top =
\begin{bmatrix}
  \lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
  \lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
\end{bmatrix}, \]

which is a GRDPG with latent vectors described by $X U$.

It is nevertheless instructive to look at a few intermediate steps. 
More specifically, the product $U I_{3, 1} U^\top$ 
yields a permutation matrix $\Pi$ with fixed points at positions $1$ and $4$ 
and a cycle of order 2 swapping positions $2$ and $3$, i.e., 
$$\Pi = U I_{3, 1} U^\top = \begin{bmatrix} 1 & 0 & 0 & 0 \\
  0 & 0 & 1 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 1
\end{bmatrix}.$$
Furthermore, since $U$ is orthonormal and $I_{3, 1}$ is diagonal, $U I_{3, 1} U^\top$
is also an eigendecomposition of $\Pi$ where the fixed
points of $\Pi$ are mapped to the eigenvectors $e_1$ and $e_4$
while the cycles of order two are mapped to the eigenvectors  
$\tfrac{1}{\sqrt{2}}(e_{2} + e_3)$ and $\tfrac{1}{\sqrt{2}}(e_{2} -
e_3)$; here $e_i$ denote the $i^\mathrm{th}$ basis vector in $\mathbb{R}^{4}$.

>>>>

\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage{parskip}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{numbers,square,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\title{Connecting the Popularity Adjusted Block Model to the Generalized Random
Dot Product Graph for Community Detection and Parameter Estimation}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
We connect two random graph models, the Popularity Adjusted Block Model
(PABM) and the Generalized Random Dot Product Graph (GRDPG),
demonstrating that a PABM is a GRDPG in which communities correspond to
certain mutually orthogonal subspaces of latent positions. This insight
leads to the construction of improved algorithms for community detection
and parameter estimation with PABM. Using established asymptotic
properties of Adjacency Spectral Embedding (ASE) for GRDPG, we derive
asymptotic properties of these algorithms, including algorithms that
rely on Sparse Subspace Clustering (SSC). We illustrate these properties
via simulation.
\end{abstract}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\newcommand{\diag}{\text{diag}}
\newcommand{\tr}{\text{Tr}}
\newcommand{\blockdiag}{\text{blockdiag}}
\newcommand{\indep}{\stackrel{\text{indep}}{\sim}}
\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
\newcommand{\Bernoulli}{\text{Bernoulli}}
\newcommand{\Betadist}{\text{Beta}}
\newcommand{\BG}{\text{BernoulliGraph}}
\newcommand{\PABM}{\text{PABM}}
\newcommand{\RDPG}{\text{RDPG}}
\newcommand{\GRDPG}{\text{GRDPG}}
\newcommand{\Multinomial}{\text{Multinomial}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem*{example}{Example}
\newpage
\spacingset{1.5} % DON'T change the spacing!

@@

Statistical analysis on graphs or networks often involves the
partitioning of a graph into disconnected subgraphs or clusters. This is
often motivated by the assumption that there exist underlying and
unobserved communities to which each vertex of the graph belongs, and
edges between pairs of vertices are determined by drawing from a
probability distribution based on the community relationships between
each pair. The goal of this analysis then is population community
detection, or the recovery of the true underlying community labels for
each vertex, up to permutation, with some additional parameter
estimation being of possible interest, assuming some underlying
probability model.

A general family of models that generate unweighted and undirected graphs is the Bernoulli Graph, which is described by an edge probability matrix $P$. Edges between each pair of vertices $i$ and $j$ is drawn as a Bernoulli trial with probability $P_{ij}$, for each $0 \leq i < j \leq n$ and $n$ is the number of vertices in the graph. Community detection methods often restrict this type of model by assigning each vertex a community label $z_1, ..., z_n$ and then setting each $P_{ij}$ as a function of the labels of vertices $i$ and $j$. Such models are called Block Models.

One popular type of Block Model is the Stochastic Block Model (SBM),
first proposed by \citet{doi:10.1080/0022250X.1971.9989788}, 
which restricts each edge probability $P_{ij}$ as only depending on the labels 
$z_i$ and $z_j$, i.e., $P_{ij} = \omega_{z_i, z_j}$. 
Various generalizations of the SBM have since been utilized, 
including the Degree-Corrected Block Model (DCBM), 
introduced by\citet{Karrer_2011}, 
which assigns an additional parameter $\theta_i$ to each vertex 
and sets $P_{ij} = \theta_i \theta_j \omega_{z_i, z_j}$. 
The Popularity Adjusted Block Model (PABM) was then introduced by
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the
DCBM to address the heterogeneity of edge probabilities within and
between communities while still maintaining distinct community
structure.

Parallel to the families of Block Models is another type of Bernoulli Graph 
called the Random Dot Product Graph (RDPG), proposed by
\citet*{10.1007/978-3-540-77004-6_11}. Under this model, each vertex of the graph
can be represented by a point in some latent space such that the edge
probability between any pair of vertices is given by their corresponding
dot product in the latent space, i.e., given a latent positions
\(x_1, ..., x_n \in \mathbb{R}^d\), the edge probability matrix is
\(P = X X^\top\) where 
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\). 
It can be shown that all Bernoulli Graphs with 
positive semidefinite edge probability matrix $P$ is a type of RDPG. 
For instance, the assortative
SBM is equivalent to a special case of the RDPG for which all vertices
of a given community share the same position in the latent space
\cite{lyzinski2014}. A similar property exists for the DCBM
\cite{lyzinski2014, rubindelanchy2017consistency}. 
Because the edge probability matrix for the PABM is not positive semidefinite, 
we instead show that it is a special case of the \emph{Generalized} Random Dot Product Graph (GRDPG) in order to analyze the PABM as a latent space model for community detection and parameter estimation.

@@

Let $G = (V, E)$ be an unweighted, undirected, and hollow graph with vertex set $V$ ($|V| = n$) and edge set $E$. $A \in \{0, 1\}^{n \times n}$ represents the adjacency matrix of $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $0$ otherwise. Because $G$ is symmetric and hollow,
$A_{ij} = A_{ji}$ and $A_{ii} = 0$ for each $i, j \in [n]$. We further restrict our analyses to Bernoulli graphs. Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ for each $1 \leq i < j \leq n$  (setting $A_{ji} = A_{ij}$ and $A_{ii} = 0$). We denote $A \sim \BG(P)$ as graph with adjacency matrix $A$ sampled from edge probability matrix $P$ in this manner. If each vertex has a hidden label in $[K]$, they are denoted as $z_1, ..., z_n$. Finally, we denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the collection of $n$ latent vectors $x_1, ..., x_n \in \mathbb{R}^d$.

\hypertarget{connecting-the-popularity-adjusted-block-model-to-the-generalized-random-dot-product-graph}{%
\section{Connecting the Popularity Adjusted Block Model to the
Generalized Random Dot Product
Graph}\label{connecting-the-popularity-adjusted-block-model-to-the-generalized-random-dot-product-graph}}

In this section, we show that the PABM is a special case of the GRDPG, i.e.,
graph $G$ drawn from the PABM can be represented by a collection of latent
vectors in Euclidean space. We further show that the latent configuration that
induces the PABM consists of orthogonal subspaces with each subspace
corresponding to a community.

\hypertarget{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}{%
\subsection{The Popularity Adjusted Block Model and the Generalized
Random Dot Product
Graph}\label{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}}

@@

Another characterization of the PABM is as follows.
Let the rows and columns of $P$ be arranged by community label
such that $n_k \times n_{\ell}$ block $P^{(k \ell)}$
describes the edge probabilities between vertices in communities
$k$ and $\ell$; note that $P^{(k \ell)} = (P^{(k \ell)})^\top$.
Now suppose that each block $P^{(k \ell)}$ can be written as the outer product of two vectors:
\begin{equation} \label{eq:pabm}
  P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top
\end{equation}
for a set of $K^2$ popularity vectors $\{\lambda^{(st)}\}_{1 \leq
  s \leq K, 1 \leq t \leq K}$ where each
$\lambda^{(st)}$ is a column vector of dimension $n_s$. Then a graph $G$
is drawn from a PABM with parameters
$\{\lambda^{(st)}\}_K$ if its adjacency matrix $A$
satisfies $A \sim \mathrm{Bernoulli}(P)$.
\end{definition}

We will use the notation \(A \sim \PABM(\{\lambda^{(kl)}\}_K)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with parameters
\(\lambda^{(kl)}\) consisting of \(K\) underlying communities.

\begin{definition}[Generalized Random Dot Product Graph]
\label{grdpg}
Let graph $G = (V, E)$ be drawn as $A \sim \BG(P)$.
If $\exists X \in \mathbb{R}^{n \times d}$ such that

@@

@@

\begin{remark}
Like the RDPG, the latent positions of a GRDPG are not unique
\cite{rubindelanchy2017statistical}.
More specifically, if $P_{ij} = x_i^\top I_{p, q} x_j$, then we also have for any
$Q \in \mathbb{O}(p, q)$,
$(Q x_i)^\top I_{p, q} (Q x_j) = x_i^\top (Q^\top I_{p, q} Q) x_j =
x_i^\top I_{p, q} x_j = P_{ij}$.
Unlike in the RDPG case, transforming the latent positions via multiplication
by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or
distances.
\end{remark}

\hypertarget{connecting-the-pabm-to-the-grdpg}{%
\subsection{Connecting the PABM to the
GRDPG}\label{connecting-the-pabm-to-the-grdpg}}
\begin{proposition}[Connecting the PABM to the GRDPG for $K = 2$]
\label{theorem1}
Let
$$X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
0 & 0 & \lambda^{(21)} & \lambda^{(22)}
\end{bmatrix} \quad \text{and} \quad
U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
0 & 1 & 0 & 0 \end{bmatrix},$$
where the $\{\lambda^{(kl)}\}$ are as defined in Definition 1.
Then $A \sim \mathrm{GRDPG}_{3, 1}(X U)$ and $B \sim \mathrm{PABM}(\{(\lambda^{(kl)}\})$ are
identically distributed.
\end{proposition}

\begin{proof}
  We can show that
\[X U I_{3, 1} U^\top X^\top =
\begin{bmatrix}
  \lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
  \lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
\end{bmatrix} \]
by straightforward matrix multiplication. It is nevertheless
instructive to look at a few intermediate steps. More specifically,
the product $U I_{3, 1} U^\top$ yields a permutation matrix $\Pi$
with fixed points at positions $1$ and $4$ and a cycle of order 2 swapping
positions $2$ and $3$, i.e., 
$$\Pi = U I_{3, 1} U^\top = \begin{bmatrix} 1 & 0 & 0 & 0 \\
  0 & 0 & 1 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 1
\end{bmatrix}.$$
Furthermore, since $U$ is orthonormal and $I_{3, 1}$ is diagonal, $U I_{3, 1} U^\top$
is also an eigendecomposition of $\Pi$ where the fixed
points of $\Pi$ are mapped to the eigenvectors $e_1$ and $e_4$
while the cycles of order two are mapped to the eigenvectors  
$\tfrac{1}{\sqrt{2}}(e_{2} + e_3)$ and $\tfrac{1}{\sqrt{2}}(e_{2} -
e_3)$; here $e_i$ denote the $i$th basis vector in $\mathbb{R}^{4}$. %  and $ eigenvalues $\pm 1$ with corresponding
% eigenvectors $(e_i \pm e_j) / \sqrt{2}$ where $i = 2, j = 3$, pair that is
% swapped.
\end{proof}
% The form for the matrix $\Pi$ used in the proof of
% Proposition~\ref{theorem1} can 
\begin{theorem}[Generalization to $K > 2$]
\label{theorem2}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\})$ be an instance of a
PABM with $K \geq 1$ blocks and latent vectors $\{\lambda^{(k \ell)}
\colon 1 \leq k \leq K, 1 \leq \ell \leq K\}$. Then there exists a block diagonal matrix
$X \in \mathbb{R}^{n \times K^2}$ defined by the $\{\lambda^{(kl)}\}$ and a $K^2 \times
K^2$ {\em fixed} orthonormal matrix $U$ such 
that $A \sim \mathrm{GRDPG}_{K (K+1) / 2, K (K-1) / 2}(XU)$.
\end{theorem}

\begin{proof}
We first define the following matrices
\begin{gather}
\label{eq:xy}
\Lambda^{(k)} = \begin{bmatrix} \lambda^{(k1)} \mid \cdots \mid \lambda^{(kK)} \end{bmatrix}
\in \mathbb{R}^{n_k \times K}, \quad
X = \text{blockdiag}(\Lambda^{(1)}, \dots, \Lambda^{(K)}) \in
\mathbb{R}^{n \times K^2}, \\
L^{(k)} = \text{blockdiag}(\lambda^{(1k)}, \dots, \lambda^{(Kk)}) \in
\mathbb{R}^{n \times K}, \quad
Y = \begin{bmatrix} L^{(1)} \mid \cdots \mid L^{(K)} \end{bmatrix} \in
\mathbb{R}^{n \times K^2}.
\end{gather}
It is then straighforward to verify that
\begin{gather*}
  XY^{\top} = \text{blockdiag}(\Lambda^{(1)}, \dots,
\Lambda^{(K)}) \begin{bmatrix} L_1^{\top} \\ \vdots \\
  L_{K}^{\top} \end{bmatrix} = \begin{bmatrix} \Lambda^{(1)}
  (L^{(1)})^{\top} \\ \vdots \\
  \Lambda^{(K)} (L^{(K)})^{\top} \end{bmatrix} \\
\Lambda^{(k)} (L^{(k)})^{\top} = \begin{bmatrix} \lambda^{(k1)}
  (\lambda^{(1k)})^{\top} \mid \dots \mid \lambda^{(kK)}
  (\lambda^{(Kk)})^{\top} \end{bmatrix} = \begin{bmatrix} P^{(k1)}
  \mid P^{(k2)} \mid \dots \mid P^{(kK)} \end{bmatrix}.
\end{gather*}
We therefore have $P = X Y^\top$. Similar to the $K = 2$ case, we also have $Y = X \Pi$ for some permutation matrix
$\Pi$ and hence $P = X \Pi X^\top$.
The permutation described by $\Pi$ has $K$ fixed points, which correspond to
$K$ eigenvalues equal to $1$ with corresponding eigenvectors $e_k$ where
$k = r (K + 1) + 1$ for $r = 0, ..., K - 1$. It also has
$\binom{K}{2} = K (K - 1) / 2$ cycles of order $2$. Each cycle corresponds to
a pair of eigenvalues $\{-1,+1\}$ and a pair of eigenvectors
$\{(e_s + e_t)/\sqrt{2},(e_s - e_t)/ \sqrt{2}\}$.
